Data Specialization for a Staged Lambda Calculus

### 2015-02-23 paper/relatedwork.tex
related work

As Jorring and Scherlis argue frequency reduction and precomputation
are common techniques for both designing algorithms and performing
compiler optimizations~\cite{JS86-staging}.  The idea behind
frequency reduction is to identify computations that are performed
multiple times and pull them ahead so that they can be performed once
and used as needed later.  The idea behind precomputation is to
identify computations that can be performed earlier, for example at
compile time if their inputs are available statically and perform them
at that earlier time. It is not difficult to see that many interesting
algorithm design techniques and compiler optimizations do exactly
frequency reduction and precomputation.  For example, dynamic
programming and loop hoisting are a frequency reduction techniques,
whereas dynamic data structures and incremental computation are
frequency reduction techniques.

Perhaps one of the most well studied example of precomputation is
partial evaluation.  Partial evaluation distinguishes between a static
(compile-time) and dynamic (run-time) stage. Given a program and the
values of its static inputs, which are the subset of the inputs that
become available at compilation time, partial evaluation generates a
specialized program by performing the computations that depend on the
statics inputs~\cite{jones96}.  The roots of partial evaluation go
back to Kleene, who in 1952 proved that a multivariate program
(function) $f(x_1, \ldots, x_{n+m})$ can be specialized for any given
$a_1 \ldots a_m$ to obtain another program $g(x_1, \ldots, x_n)$ such
that $f(a_1, \ldots, a_m, y_1, \ldots, y_n) = g(y_1, \ldots,
y_n)$~\cite{Kleene52}.  Kleene did not intend to improve efficiency,
however.  Lombardi appears to be the first person to use the term in
his work on incremental computation~\cine{Lombardi67}.  Since then,
there has a lot of work on partial evaluation, the only broad-strokes
of which we can mention in this here.  We refer the interested reader
to the excellent book by Jones, Gommard, and Sestoft for a
comprehesive discussion of work until early 90's~\cite{JGS93}.

At a high level, earlier approaches to partial evaluation can be
viewed as operating in two stages: binding time analysis and program
specialization.  For a multivariate program, e.g., $f(x_s,x_d)$, with
clearly marked static and dynamic arguments, binding-time analysis
identifies all the parts of the program that can be computed by the
knowledge of static arguments. Using this information and the static
arguments themselves, program specialization specializes the program
to a partially evaluated program that can be used with many different
dynamic arguments.  This approach has been applied to construct
partial evaluators for a number of languages such as 
Scheme~\cite{OB91-Similix,Consel88-Schism}, considering for the most
part first-order subsets.

Subsequence experience with binding time analysis based approach
showed that it can be difficult to control, leading to specialized
programs whose performance can be difficult to predict. This led to
investigations based on type systems for making explicit in the
program the stage of each
computation~\cite{GJ91-lambda,NN92-twolevel}.  Davies~\cite{Davies96}
presented a logical construction of binding-time type systems by
deriving a type system via the Curry-Howard isomorphism applied to
temporal logic.  In subsequence work, Davies and Pfenning proposed a
new type system for staged computation based on a particular fragment
of modal logic~\cite{DP01-modal}.  These later approaches can be
viewed more generally meta programming, where evaluating a program at
a certain stage with its inputs at that stage gives another program to
be executed at the next stage (e.g.,~\cite{JS86-staging}). The work on
MetaML scaled type-based ideas to a full-scale language by developing
statically typed programming language based on ML that enables the
programmer to express programs with multiple
stages~\cite{Taha97,taha-thesis-99}.  MetaML's type system is similar
to Davies~\cite{Davies96} but extends it in several important ways.
Nanevksi and Pfenning further extended the these techniques by
allowing free variable to occur within staged
computations~\cite{NP05-nn}.

>>>>

As Jorring and Scherlis argue frequency reduction and precomputation
are common techniques for both designing algorithms and performing
compiler optimizations~\cite{JS86-staging}.  The idea behind frequency
reduction is to identify computations that are performed multiple
times and pull them ahead so that they can be performed once and used
later as needed.  The idea behind precomputation is to identify
computations that can be performed earlier, for example at compile
time if their inputs are available statically and perform them at that
earlier time.  Many interesting algorithm design techniques and
compiler optimizations perform frequency reduction and precomputation.
For example, dynamic programming and loop hoisting are a frequency
reduction techniques, whereas dynamic data structures and incremental
computation are precomputation techniques.

Perhaps one of the most well studied example of precomputation is
partial evaluation.  Partial evaluation distinguishes between a static
(compile-time) and dynamic (run-time) stage. Given a program and the
values of its static inputs, which are the subset of the inputs that
become available at compilation time, partial evaluation generates a
specialized program by performing the computations that depend on the
statics inputs~\cite{jones96}.  The roots of partial evaluation go
back to Kleene, who in 1952 proved that a multivariate program
(function) $f(x_1, \ldots, x_{n+m})$ can be specialized for any given
$a_1 \ldots a_m$ to obtain another program $g(x_1, \ldots, x_n)$ such
that $f(a_1, \ldots, a_m, y_1, \ldots, y_n) = g(y_1, \ldots,
y_n)$~\cite{Kleene52}.  Kleene did not intend to improve efficiency,
however.  Lombardi appears to be the first person to use the term in
his work on incremental computation~\cite{Lombardi67}.  Since then,
there has a lot of work on partial evaluation, only broad-strokes of
which we can mention in this here.  We refer the interested reader to
the excellent book by Jones, Gommard, and Sestoft for a comprehesive
discussion of work until early 90's~\cite{JGS93}.

At a high level, earlier approaches to partial evaluation can be
viewed as operating in two stages: binding time analysis and program
specialization.  For a multivariate program with clearly marked static
and dynamic arguments, binding-time analysis identifies all the parts
of the program that can be computed by the knowledge of static
arguments. Using this information and the static arguments themselves,
program specialization specializes the program to a partially
evaluated program that can be used with many different dynamic
arguments.  This approach has been applied to construct partial
evaluators for a number of languages such as
Scheme~\cite{OB91-Similix,Consel88-Schism}, considering for the most
part first-order subsets.

Experience with binding time analysis showed that it can be difficult
to control, leading to programs whose performance can be difficult to
predict. This led to investigations based on type systems for making
explicit in the program the stage of each
computation~\cite{GJ91-lambda,NN92-twolevel} and writing ``meta
programs'' that when evaluated at a stage, give another program to be
evaluated at the next stage. Jorring and Scherlis describe this idea
meta programming in their paper (~\cite{JS86-staging}) as one of the
three possible approaches to taking advantage of
staging. Davies~\cite{Davies96} presented a logical construction of
binding-time type systems by deriving a type system via the
Curry-Howard isomorphism applied to temporal logic.  Davies and
Pfenning proposed a new type system for staged computation based on a
particular fragment of modal logic~\cite{DP01-modal}. The work on
MetaML extended type-based techniques to a full-scale language by
developing statically typed programming language based on ML that
enables the programmer to express programs with multiple
stages~\cite{Taha97,taha-thesis-99}.  MetaML's type system is similar
to Davies~\cite{Davies96} but extends it in several important ways.
Nanevksi and Pfenning further extended the these techniques by
allowing free variable to occur within staged
computations~\cite{NP05-nn}.
